\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-12-01 smurf_paper.tex
sky noise and clarification of whitened PSF

Data from the fully commissioned instrument, observation 38 on
20111112, are shown in Fig.~\ref{fig:bolos_mix}b. Having solved the
fridge oscillation problem, these data no longer exhibit a correlation
with the mixing chamber (however, note that variations are still seen
in the mixing chamber signal; such variations do not necessarily
reflect changes in the focal plane temperature). There is minimal
correlated low-frequency noise in these 450\,\micron\ data, resulting
in little difference between the raw and common-mode subtracted
data. The 850\,\micron\ channel, however, exhibits a more significant
signal that is obviously correlated with the telescope motion. It is
also correlated amongst many of the detectors, and common-mode removal
corrects it to a large extent. Part of the reason that this is seen
here, and not in the S2SRO data set shown, is that the amplitude of
the scan pattern is larger. Most of this ``scan-synchronous'' noise is
attributed to magnetic field pickup, as described in
Section~\ref{sec:magpickup}. Also similar to the S2SRO data, an
anti-correlation between the opacity and elevation is apparent.

@@

An additional noise source that is significant in only a small subset
of the data (and more so during the S2SRO period) is magnetic field
pickup. Since the bolometer signals are ultimately detected through
the amplification of magnetic fields, any additional changing fields
within the instrument will also be detected.

@@

@@

A method that has been used to remove correlated noise as part of the
map-making procedure for Bolocam and AzTEC is Principal Component
Analysis \citep[PCA,][]{laurent2005,scott2008,perera2008}. Here we use
PCA to explore correlated signals in \scuba\ data.

@@

The majority of the correlated signal at both wavelengths in
Fig.~\ref{fig:pca} is produced by Component~1, in which the
eigenvector time-series exhibits a roughly periodic signal that
resembles the scan pattern in Fig.~\ref{fig:bolos_mix}b. While this
correlated signal appears at both wavelengths, referring to the maps
of eigenvalues, they are considerably stronger at 850\,\micron\
(consistent with the visual appearance of the bolometer signals in
Fig.~\ref{fig:pca}). Also note that the PSD for theis eigenvector
exhibit nearly a pure $1/f$ signature with minimal line features, or a
white-noise plateau (suggesting that these are high-\snr\ measurements
of a purely low-frequency drift). Component~3 is weaker, yet also
shows evidence of the scan-synchronous noise, combined with
high-frequency line features that appear primarily in columns 26, 29
and 30 of the 450\,\micron\ subarray. Unlike Component~1, it is
generally more dominant at 450\,\micron.

@@

Why are there two components that seem to be dominated by
scan-synchronous noise (which is presumably a mixture of
magnetic-field pickup and atmospheric noise)? First, the
smoothly-varying gradients in the eigenvalue maps (primarily at
850\,\micron) suggests that this noise has a different response across
the focal plane; in this case the PCA has identified two orthogonal
shapes that, when mixed in different quantities, can reproduce most of
the signal in each bolometer. It is also likely that the atmospheric
contribution is stronger at 450\,\micron\ (i.e., predominantly
Component~3).

Like Componen~1, Component~2 is a purely $1/f$ signal, with an
amplitude comparable to Component~3, that appears almost exclusively
at 850\,\micron. Since it is not obviously correlated with the
telescope motion, yet has an amplitude that varies smoothly across the
850\,\micron\ subarray, it may represent the PCAs attempt to reproduce
some non-linear response to magnetic-field and/or atmospheric pickup

The remaining, weaker, Components~4--9 share some common
features. They all have some degree of $1/f$ noise, suggesting further
residual noise related to the telescope motion and/or other slow
baseline drifts. All of them also exhibit line features in their
spectra, which in many cases appear at random amplitudes along columns
(e.g. columns 15, and 26--30 at 450\,\micron\ in Component~4), or in a
sampling of isolated bolometers (e.g. colummn 13, row 19 at
850\,\micron\ in Component~6). The signals that appear along columns
is probably related to magnetic-field pickup. Other sources of lines
are thought to be aliased high-frequency noise, and also the 60\,Hz
mains which appear in some bolometers.

This example illustrates some of the types of correlated signals and
patterns that PCA can identify in \scuba\ data. While there are
typically one or two strong signal components detected, in general,
the details can vary significantly from data set to data set, and the
lengths of the time-series analysed. The reason for this is that many
of the noise sources are not stationary in time (e.g.,
scan-synchronous noise which obviously depends on the scan pattern,
and the tuning of the subarrays). It should also be clear from this
example that while PCA offers some helpful insight into the various
sources of noise, it does not necessarily identify clean patterns that
can easily be modelled. For example, while there are clearly
correlation patterns along columns as expected from the common
amplification chain, the intensities of these high-frequency signals
appear random in the eigenvalue maps, and a simple common-mode signal
estimated from the data for each column would not remove
it. Regardless, the dominant noise, in all cases, is at
low-frequencies. This conclusion is central to our data reduction
strategy described in Section~\ref{sec:algorithm}.

\subsection{Overall performance and mapping speed}

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{lockman_psf}
\caption{Slice through the angular response to an ideal Gaussian point
  source (solid line) along the R.A. axis following map-making using
  the blank-field processing configuration with SMURF (dotted line),
  and upon application of the whitening filter (dashed line). The
  ``map filtered'' response is produced by adding the ideal Gaussian
  to the real data near the centre of the Lockman Hole map, and
  re-reducing the data. The resulting dotted line gives the expected
  shape of a source in panel (a) from
  Fig.~\ref{fig:lockman_maps}. Applying the whitening filter
  (reciprocal of the solid orange line in
  Fig.~\ref{fig:lockman_pspec}) to the dotted line gives the whitened
  line (dashed) , which is the expected shape of a source in panel (b)
  from Fig.~\ref{fig:lockman_maps}. Finally, cross-correlating the
  whitened map with this whitened PSF is an effective ``matched
  filter'' for identifying point-like sources, and this smoothed map
  is shown in panel (c) of Fig.~\ref{fig:lockman_maps}.}
\label{fig:lockman_psf}
\end{figure}

@@

The angular power spectrum of the whitened signal map is shown with a
solid line in Fig.~\ref{fig:lockman_pspec}. At low angular frequencies
($\lsim 0.5$\,arcsec$^{-1}$) there is significantly less power than
for the raw map. However, it also clearly has some power in excess of
the white noise level. In theory, this residual signal is produced by
astronomical sources, although its origin cannot be determined from
this plot alone; nor are astronomical sources readily visible in the
map (Fig.~\ref{fig:lockman_maps}b). To identify sources, we next apply
a matched filter to the whitened maps.

For blind, high-redshift surveys, individual sources are expected to
be un-resolved by the \scuba\ 7.5--14.5\,arcsec FWHM beams. Under this
assumption, and also assuming that the map noise is white,
cross-correlation between the map and the known PSF yields the
maximum-likelihood flux density of supposed point-sources centred over
every location in the resulting map \citep[an extremely well-known
result throughout astronomy, see][]{stetson1987}. Peak identification
in such smoothed maps have been used extensively in the submillimetre
community, as both an efficient source-detection and photometric
measurement strategy. In terms of the angular power spectrum, a
matched filter may also be thought of as an optimal low-pass filter,
that suppresses noise on scales smaller (and therefore at higher
angular frequencies) than the beam. For the case at hand, the
effective PSF for the whitened maps is given by the dotted curve in
Fig.~\ref{fig:lockman_psf}. Maps smoothed by this shape are shown for
the whitened signal, and jackknife maps in
Figs.~\ref{fig:lockman_maps}c and f, respectively. Note that these
images are plotted in \snr\ units, where the smooth noise maps have
been calculated by propagating the original noise maps output by SMURF
both through the whitening and matched filters (each of which are
linear operations).

>>>>

Data from the fully commissioned instrument, observation 38 on
20111112, are shown in Fig.~\ref{fig:bolos_mix}b. Having solved the
fridge oscillation problem, these data no longer exhibit a correlation
with the mixing chamber (however, note that variations are still seen
in the mixing chamber signal; such variations do not necessarily
reflect changes in the focal plane temperature). There is minimal
correlated low-frequency noise in these 450\,\micron\ data, resulting
in little difference between the raw and common-mode subtracted
data. The 850\,\micron\ channel, however, exhibits a more significant
signal that is obviously correlated with the telescope motion. It is
also correlated amongst many of the detectors, and common-mode removal
corrects it to a large extent. Part of the reason that this is seen
here, and not in the S2SRO data set shown, is that the amplitude of
the scan pattern is larger. This ``scan-synchronous'' noise is
attributed to a combination of magnetic field pickup, as described in
Section~\ref{sec:magpickup}, and sky brightness variations due to
changes in elevation (not to be confused with underlying changes in
the atmosphere; a homogeneous atmosphere will appear brighter with
increasing airmass). Also similar to the S2SRO data, an
anti-correlation between the opacity and elevation is apparent.

@@

An additional noise source that is significant primarily for wide-area
scans (and more so during the S2SRO period) is magnetic field
pickup. Since the bolometer signals are ultimately detected through
the amplification of magnetic fields, any additional changing fields
within the instrument will also be detected.

@@

%-------------------------------------------------
\subsection{Sky noise}
\label{sec:skynoise}
%-------------------------------------------------

\begin{figure}
\centering
\includegraphics[width=\linewidth]{skynoise}
\caption{Illustration of the dominance of atmospheric variations on
  long-timescales using 35\,min. of data from the fully-commissioned
  instrument, observation 18 on 20110324. Red and blue lines show the
  850 and 450\,micron\ common-mode signals, corresponding to the left-
  and right- vertical axes, respectively. The black line shows the
  line-of-sight opacity from the WVM as a proxy for atmospheric
  emission (arbitrary scale). Bolometer baseline drifts on timescales
  $\gsim$100\,s are clearly correlated with the atmosphere in this
  case.}
\label{fig:skynoise}
\end{figure}

The data shown in Figs.~\ref{fig:bolos_mix} and \ref{fig:pspec} are
dominated at low-frequencies ($\lsim$2\,Hz) by scan-synchronous noise,
rathern than underlying variations in the atmosphere, or ``sky
noise''. However, only $\sim$100\,s of data were analysed. In
Fig.~\ref{fig:skynoise}, an order-of-magnitude longer
(fully-commissioned) data set is plotted from observation 18 on
20110324. The common-mode signals from the 450 and 850\,\micron\ focal
planes are compared with the line-of-sight opacity at 183\,GHz
(1.6\,mm) from the WVM \citep[linear relationships between this
opacity, and those at 450 and 850\,micron\ are given
in][]{dempsey2012}. Unlike the earlier plots, it is clear that the
bolometer baseline drifts on timescales $\gsim$100\,s are tightly
correlated with the atmospheric opacity (especially at 850\,\micron),
which serves as a proxy for atmospheric emission. The fact that the
correlation between the WVM and the 450\,micron\ bolometers is weaker
either reflects an underlying different in the atmosphere at higher
frequencies, or the presence of some other significant low-frequency
noise source.

While the relative importance of sky noise at low-frequencies is
highly time- and scan-pattern dependent, in general it does not have a
major impact on map-making. Upon common-mode removal (which is an
integral part of our map-making strategy) uncorrelated noise remains
(e.g., solid lines in Fig~\ref{fig:pspec}). As we will see in
Section~\ref{sec:pca} and \ref{sec:experimental}, this remaining noise
does not have a smoothly-varying correlation pattern across the focal
plane one might expect if it were due to resolved ``clouds'' of
atmospheric emission. A high-pass filter above $\sim$0.1--1.0\,Hz is
usually employed remove it (Section~\ref{sec:components}), which will
also remove any sky noise that is not already subtracted with the
common-mode.

@@

A method that has been used to remove correlated noise as part of the
map-making procedure for Bolocam and AzTEC is Principal Component
Analysis \citep[PCA,][]{laurent2005,scott2008,perera2008}. Here we use
PCA to further explore correlated signals in \scuba\ data.

@@

The majority of the correlated signal at both wavelengths in
Fig.~\ref{fig:pca} is produced by Component~1, in which the
eigenvector time-series exhibits a roughly periodic signal that
resembles the scan pattern in Fig.~\ref{fig:bolos_mix}b. While this
correlated signal appears at both wavelengths, referring to the maps
of eigenvalues, they are considerably stronger at 850\,\micron\
(consistent with the visual appearance of the bolometer signals in
Fig.~\ref{fig:pca}). Also note that the PSD for this eigenvector
exhibits nearly a pure $1/f$ signature with minimal line features, and
no white-noise plateau (suggesting that these are high-\snr\
measurements of a purely low-frequency drift). Component~3 is weaker,
yet also shows evidence of the scan-synchronous noise, combined with
high-frequency line features that appear primarily in columns 26, 29
and 30 of the 450\,\micron\ subarray. Unlike Component~1, it is
generally more dominant at 450\,\micron.

@@

Why are there two components that seem to be dominated by
scan-synchronous noise (which is presumably a mixture of
magnetic-field pickup and brightness variations in the atmosphere)?
First, the smoothly-varying gradients in the eigenvalue maps
(primarily at 850\,\micron) suggests that this noise has a different
response across the focal plane; in this case the PCA has identified
two orthogonal shapes that, when mixed in different quantities, can
reproduce most of the signal in each bolometer. It is also likely that
the atmospheric contribution is stronger at 450\,\micron\ (i.e.,
predominantly Component~3).

Like Component~1, Component~2 is a purely $1/f$ signal, with an
amplitude comparable to Component~3, that appears almost exclusively
at 850\,\micron. Since it is not obviously correlated with the
telescope motion, yet has an amplitude that varies smoothly across the
850\,\micron\ subarray, it may represent the PCAs attempt to reproduce
some non-linear response to magnetic-field and/or atmospheric pickup

The remaining, weaker, Components~4--9 share some common
features. They all have some degree of $1/f$ noise, suggesting further
residual noise related to the telescope motion and/or other slow
baseline drifts. All of them also exhibit line features in their
spectra, which in many cases appear at random amplitudes along columns
(e.g. columns 15, and 26--30 at 450\,\micron\ in Component~4), or in a
sampling of isolated bolometers (e.g. colummn 13, row 19 at
850\,\micron\ in Component~6). The signals that appear along columns
are probably related to magnetic-field pickup. Other sources of lines
are thought to be aliased high-frequency noise, and also the 60\,Hz
mains which appear in some bolometers.

This example illustrates some of the types of correlated signals and
patterns that PCA can identify in \scuba\ data. While there are
typically one or two strong signal components detected, in general,
the details can vary significantly from data set to data set, and the
lengths of the time-series analysed. The reason for this is that many
of the noise sources are not stationary in time (e.g.,
scan-synchronous noise which obviously depends on the scan pattern,
and the tuning of the subarrays). It should also be clear from this
example that while PCA offers some helpful insight into the various
sources of noise, it does not necessarily identify clean patterns that
can easily be modelled. For example, while there are clearly
correlation patterns along columns as expected from the common
amplification chain, the intensities of these high-frequency signals
appear random in the eigenvalue maps, and a simple common-mode signal
estimated from the data for each column would not remove
it. Regardless, the dominant noise in excess of the fundamental white
noise limit, in all cases, is at low-frequencies. This conclusion is
central to our data reduction strategy described in
Section~\ref{sec:algorithm}.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{lockman_psf}
\caption{Slice through the angular response to an ideal Gaussian point
  source (solid line) along the R.A. axis following map-making using
  the blank-field processing configuration with SMURF (dotted line),
  and upon application of the whitening filter (dashed line). The
  ``map filtered'' response is produced by adding the ideal Gaussian
  to the real data near the centre of the Lockman Hole map, and
  re-reducing the data. The resulting dotted line gives the expected
  shape of a source in panel (a) from
  Fig.~\ref{fig:lockman_maps}. Applying the whitening filter
  (reciprocal of the solid orange line in
  Fig.~\ref{fig:lockman_pspec}) to the dotted line gives the whitened
  line (dashed), which is the expected shape of a source in panel (b)
  from Fig.~\ref{fig:lockman_maps}. Finally, cross-correlating the
  whitened map with this whitened PSF is an effective ``matched
  filter'' for identifying point-like sources, and this smoothed map
  is shown in panel (c) of Fig.~\ref{fig:lockman_maps}.}
\label{fig:lockman_psf}
\end{figure}

@@

The angular power spectrum of the whitened signal map is shown with a
solid line in Fig.~\ref{fig:lockman_pspec}. At low angular frequencies
($\lsim 0.5$\,arcsec$^{-1}$) there is significantly less power than
for the raw map. However, it also clearly has some power in excess of
the white noise level. In theory, this residual signal is produced by
astronomical sources, although its origin cannot be determined from
this plot alone; nor are astronomical sources readily visible in the
map (Fig.~\ref{fig:lockman_maps}b). To identify sources, we next apply
a ``matched filter'' to the whitened maps.

For blind, high-redshift surveys, individual sources are expected to
be un-resolved by the \scuba\ 7.5--14.5\,arcsec FWHM beams. Under this
assumption, and also assuming that the map noise is white,
cross-correlation between the map and the known PSF, or matched
filtering, yields the maximum-likelihood flux density of supposed
point-sources centred over every location in the resulting map
\citep[an extremely well-known result throughout astronomy,
see][]{stetson1987}. Peak identification in such smoothed maps have
been used extensively in the submillimetre community, as both an
efficient source-detection and photometric measurement strategy. In
terms of the angular power spectrum, a matched filter may also be
thought of as an optimal low-pass filter, that suppresses noise on
scales smaller (and therefore at higher angular frequencies) than the
beam. For the case at hand, we may use this cross-correlation
technique since the map has been whitened. However, we must first
establish the effective shape of point sources in this map due both to
map-making itself, and the whitening filter. We determine the effect
of map-making by adding a synthetic (and high-\snr) point source to
the real data, and measure its resulting shape in the map (solid and
dotted lines in Fig.~\ref{fig:lockman_psf}, respectively). Next, the
Fourier transform of the map filtered PSF is divided by the jackknife
noise power spectrum to calculate the final whitened PSF (dashed line
in Fig.~\ref{fig:lockmand_psf}). Maps smoothed by this shape are shown
for the whitened signal, and jackknife maps in
Figs.~\ref{fig:lockman_maps}c and f, respectively. Note that these
images are plotted in \snr\ units, where the smooth noise maps have
been calculated by propagating the original noise maps output by SMURF
both through the whitening and matched filters (each of which are
linear operations).
